Difference Scores: Paired Samples T-test

### 2024-03-07 examples/_1a_Examples_DifferenceScores_PairedTtests.qmd
codespace updates

# Create a list of required packages
packages_required <- c("tidyverse","easystats","lme4","gtsummary","report","broom","gridExtra")

@@

```
:::

@@

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
                      comment=NA, cache=T, code_folding=T,
                      R.options=list(width=220, digits = 3),
                      fig.align='center', 
                      out.width='75%', fig.asp=.75)
```
:::
:::

@@

::: blue
**This code reads in and shows the data to be used in the current
example**

```{r read-data}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE
#| cache: FALSE

@@

```
:::

----

# Authenticate with Google Drive

----

@@

descriptives_table <- df %>%
  select(eventname, sex, race.4level, hisp, anthroweightcalc) %>%
  mutate(eventname = factor(eventname, labels = c("Baseline", "Year 1","Year 2","Year 3","Year 4"))) %>%
  mutate(sex = factor(sex, labels = c("Female", "Male"))) %>%
  tbl_summary(
    by = eventname,
    missing = "no",
    label = list(sex ~ "Sex", race.4level ~ "Race", hisp ~ "Hispanic", 
                 anthroweightcalc ~ "Weight"),
    statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
  ) %>%
modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
  bold_labels() %>%
  italicize_levels() %>%
  modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

@@

>>>>

# Create a list of required packages
packages_required <- c("tidyverse", "easystats", "lme4", "gtsummary", "report", "broom", "gridExtra")

@@

:::

@@

knitr::opts_chunk$set(
    echo = T, message = F, warning = F, error = F,
    comment = NA, cache = T, code_folding = T,
    R.options = list(width = 220, digits = 3),
    fig.align = "center",
    out.width = "75%", fig.asp = .75
)
```

:::

@@

# Authenticate with Google Drive
drive_auth()

:::

<!-- ::: blue
**This code reads in and shows the data to be used in the current
example** -->

<!-- ```{r read-data}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE
#| cache: FALSE

@@

```
::: -->

@@

df <- my_data

descriptives_table <- df %>%
    select(eventname, sex, race.4level, hisp, anthroweightcalc) %>%
    mutate(eventname = factor(eventname, labels = c("Baseline", "Year 1", "Year 2", "Year 3", "Year 4"))) %>%
    mutate(sex = factor(sex, labels = c("Female", "Male"))) %>%
    tbl_summary(
        by = eventname,
        missing = "no",
        label = list(
            sex ~ "Sex", race.4level ~ "Race", hisp ~ "Hispanic",
            anthroweightcalc ~ "Weight"
        ),
        statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
    ) %>%
    modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
    bold_labels() %>%
    italicize_levels() %>%
    modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

@@

**Descriptive Statistics for the Difference Score**
```{r model-checking}
# Compute statistical summaries for the difference score variable
lapply(difference_and_summary_list, function(item) {
    print(item$summary)
})

gaze <- df %>%
  mutate(diff = baseline_event - followup_event) %>%
  mutate(group_resid = diff - mean(diff))

gaze <- weight_diff_data %>%
  mutate(diff1 = baseline_year_1_arm_1 - 1_year_follow_up_y_arm_1) %>%
  mutate(group_resid = diff - mean(diff))

dff.y

qqPlot(gaze_residual$group_resid)

shapiro.test(gaze_residual$group_resid)

This summary of the difference score variable indicates that the differences range from a decrease of `r min(weight_diff_data$diff, na.rm = TRUE)` units to an increase of `r max(weight_diff_data$diff, na.rm = TRUE)` units. The median difference is `r median(weight_diff_data$diff, na.rm = TRUE)` units, and the average difference is approximately `r mean(weight_diff_data$diff, digits = 2, na.rm = TRUE)` units. There are `r sum(is.na(weight_diff_data$diff))` missing values in this difference score dataset.

**Descriptive Statistics for the weight Variable**
```{r}
## Summary statistics
# Compute summary statistics for weight by eventname
library(dplyr)
summary <- df %>%
    group_by(eventname) %>%
    get_summary_stats(anthroweightcalc, type = "mean_sd")

data.frame(summary)
```

The summary statistics provide insights into the weight variable across different `eventname` categories. At baseline, the average weight is approximately `r summary$mean[summary$eventname == "baseline_year_1_arm_1"]` units with a standard deviation of `r summary$sd[summary$eventname == "baseline_year_1_arm_1"]`. Over the years, there’s a noticeable increase in average weight: by the first year, it’s about `r summary$mean[summary$eventname == "1_year_follow_up_y_arm_1"]` units, and it continues to rise, reaching approximately `r summary$mean[summary$eventname == "4_year_follow_up_y_arm_1"]` units by the fourth year. The standard deviation remains relatively consistent over the years, suggesting similar variability in weight measurements across these time points.

:::

### Build Model: Step 2 {.tabset .tabset-fade .tabset-pills}

::: blue

A one-sample t-test is computed to examine whether the average difference score is different than zero (indicating change).

**STEP 2: Conduct t-test on Difference Score**

```{r build-model-ttest-2}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

# Extract the difference scores for weight from the computed list
diff_data <- difference_and_summary_list[[1]]$data$diff

# Perform a one-sample t-test on the difference scores for weight
test_result <- t.test(diff_data, mu = 0, na.rm = TRUE)

test_result

gaze_test <- t.test(anthroweightcalc ~ eventname, 
                    paired = TRUE, 
                    data = df) %>% 
  tidy()

This one-sample t-test evaluates whether the average difference score deviates significantly from 0. The model output provides:

i. a t-statistic of `r test_result$statistic`;
ii. degrees of freedom of `r test_result$parameter`;
iii. a p-value of `r test_result$p.value`;
iv. a mean estimate of `r test_result$estimate["mean of x"]`;
v. 95% confidence intervals from `r test_result$conf.int[1]` to `r test_result$conf.int[2]`.

v. 95% confidence intervals from `r round(test_result$conf.int[1], 3)` to `r round(test_result$conf.int[2], 3)`.

With a t-value of `r test_result$statistic` and a p-value of `r test_result$p.value`, we can determine the significance of the difference. The average difference score across the sample is approximately `r round(test_result$estimate["mean of x"], 2)`. Furthermore, we can be 95% confident that the true mean difference score in the population lies between approximately `r round(test_result$conf.int[1], 2)` and `r round(test_result$conf.int[2], 2)`. This indicates the degree to which there's a change in weight from Baseline to Year 1.

:::

### Model Plots

::: blue
```{r scatterplot}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| code-summary: testing

# Filter data for Baseline
baseline_data <- df %>% filter(eventname == "Baseline")

# Filter data for Year_1 and join with Baseline data
joined_data <- baseline_data %>%
    left_join(df %>% filter(eventname == "Year_1"), by = "src_subject_id", suffix = c("_baseline", "_year1"))

# Create scatterplot
scatterplot <- ggplot(df, aes(x = anthroweightcalc, y = diff)) +
    geom_point(aes(color = sex), alpha = 0.5) + # Setting alpha to ensure points are visible
    geom_smooth(method = "lm", color = "red") + # Adding linear regression line
    labs(
        title = "Scatterplot of anthroweightcalc vs. Difference Score",
        x = "anthroweightcalc",
        y = "Difference Score"
    ) +
    theme_minimal()

scatterplot

The scatterplot visually depicts the relationship between individuals' weights at baseline and their weights at Year_1. Each point on the plot represents an individual, with their baseline weight plotted on the x-axis and their Year_1 weight on the y-axis. A noticeable positive linear trend can be observed, as highlighted by the blue regression line, indicating that those who were taller at baseline generally remained taller at Year_1. The strength and direction of this relationship can be further quantified with correlation coefficients, but visually, the data suggests a strong positive association between baseline and Year_1 weights.
